Weighted Model Counting with Conditional Probabilities

### 2020-07-18 paper.tex
Prepare for big experiments
Proof that the encoding is correct

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{complexity}
\usepackage{soul}

@@

\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\nWMC}{NWMC}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\End}{End}

@@

@@

@@

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{hyperref}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[backgroundcolor=lightgray]{todonotes}
\usepackage{complexity}
\usepackage{soul}
\usepackage[ruled,vlined]{algorithm2e}

@@

\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\nWMC}{NWMC}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\im}{im}

@@

\item Prove that we have an algebra of functions with one additional operation.
\item The function $\phi$ created by the algorithm can be seen as a measure on
  the BA $2^U$. Can I at least show that this BA has fewer elements as a measure
  of compactness w.r.t. the measure?
\item Note that we're going to use logical notation for BAs. Give an example of
  the two notations.
\item Show how the $\phi$ from the algorithm can be encoded in the typical
  DIMACS CNF format.

@@

\item We assume that all variables in the Bayesian network have at least two
  values.

@@

\section{Encoding Bayesian networks}

Let $V$ denote the set of random variables in a Bayesian network. For any random
variable $X \in V$, let $\mathrm{pa}(X)$ denote the set of parents of $X$ and
$\im X$ denote the set of possible values.

\begin{definition}[Indicator variables]
  Let $X \in V$ be a random variable. If $X$ is binary (i.e., $|\im X| = 2$), we
  can arbitrary identify one of the values as $1$ and the other one as $0$ (i.e,
  $\im X \cong \{ 0, 1 \}$). Then $X$ can be represented by a single
  \emph{indicator variable} $\lambda_{X=1}$. If we interpret
  $2^{\{\lambda_{X=1}\}}$ as a Boolean algebra, we can let $\lambda_{X=0} =
  \neg\lambda_{X=1}$ in the algebraic notation, or, equivalently, $\lambda_{X=0}
  = \emptyset \in 2^{\{\lambda_{X=1}\}}$ in the set-theoretic notation. On the
  other hand, if $X$ is not binary, we represent $X$ with $|\im X|$ indicator
  variables, one for each value. We let
  \[
    E(X) = \begin{cases}
      \{ \lambda_{X=1} \} & \text{if } |\im X| = 2 \\
      \{ \lambda_{X=x} \mid x \in \im X \} & \text{otherwise.}
    \end{cases}
  \]
  denote the set of indicator variables for $X$ and
  \[
    E^*(X) = E(X) \cup \bigcup_{Y \in \mathrm{pa}(X)} E(Y).
  \]
  denote the set of indicator variables for $X$ and its parents in the Bayesian
  network. Finally, let
  \[
    U = \bigcup_{X \in V} E(X)
  \]
  denote the set of all indicator variables for all random variables in the
  Bayesian network.
\end{definition}

\begin{definition}[Operations on functions]
  Let $A\colon 2^X \to [0, 1]$ and $B\colon 2^Y \to [0, 1]$ be Boolean
  functions, $\alpha \in [0, 1]$, and $x \in X$. We define the following
  operations:
  \begin{description}
  \item[Addition:] $A+B$ is a function $A+B\colon 2^{X \cup Y} \to [0, 1]$ such
    that
    \[
      (A+B)(\tau) = A(\tau \cap X) + B(\tau \cap Y)
    \]
    for all $\tau \in 2^{X \cup Y}$.
  \item[Inverse:] $\overline{A}$ is a function $\overline{A}\colon 2^X \to [0,
    1]$ such that
    \[
      \overline{A}(\tau) = 1 - A(\tau)
    \]
    for all $\tau \in 2^X$.
  \item[Multiplication:] $A \cdot B$ is a function $A \cdot B\colon 2^{X \cup Y}
    \to [0, 1]$ such that
    \[
      (A \cdot B)(\tau) = A(\tau \cap X) \cdot B(\tau \cap Y)
    \]
    for all $\tau \in 2^{X \cup Y}$.
  \item[Scalar multiplication:] $\alpha A$ is a function $\alpha A\colon 2^X \to
    [0, 1]$ such that
    \[
      (\alpha A)(\tau) = \alpha \cdot A(\tau)
    \]
    for all $\tau \in 2^X$.
  \item[Projection:] $\exists_xA$ is a function $\exists_xA\colon 2^{X \setminus
      \{ x \}} \to [0, 1]$ such that
    \[
      (\exists_xA)(\tau) = A(\tau) + A(\tau \cup \{ x \})
    \]
    for all $\tau \in 2^{X \setminus \{x \}}$.
  \end{description}
\end{definition}
\begin{remark}
  Note that $A + \overline{A} = 1$.
\end{remark}

\begin{definition}[Special functions]
  \phantom{}
  \begin{itemize}
  \item unit $1\colon 2^\emptyset \to [0, 1]$, $1(\tau) = 1$.
  \item zero $0\colon 2^\emptyset \to [0, 1]$, $0(\tau) = 0$.
  \item constant $[a]\colon 2^{\{a\}} \to [0, 1]$,
    \[
      [a](\tau) =
      \begin{cases}
        1 & \text{if } a \in \tau \\
        0 & \text{if } a \not\in \tau.
      \end{cases}
    \]
  \end{itemize}
\end{definition}

\begin{algorithm}
  $\phi \gets 1$\;
  \For{$X \in V$}{
    \textit{let} $\mathrm{pa}(X) = \{ Y_1, \dots, Y_n \}$\;
    \eIf{$|\im X| = 2$}{
      \For{$(y_1, \dots, y_n) \in \prod_{i = 1}^n \im Y_i$}{
        $p_1 \gets \Pr(X = 1 \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $p_0 \gets \Pr(X \ne 1 \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $\phi \gets \phi \cdot \left( p_1[\lambda_{X=1}] \cdot \prod_{i=1}^n
          [\lambda_{Y_i=y_i}] + p_0 \overline{[\lambda_{X=1}]} \cdot
          \prod_{i=1}^n [\lambda_{Y_i=y_i}] \right)$\;
      }
    }{
      \textit{let} $\im X = \{ x_1, \dots, x_m \}$\;
      $\phi \gets \phi \cdot \left( \sum_{i=1}^m [\lambda_{X = x_i}] \right)
      \cdot \prod_{i=1}^m \prod_{j=i+1}^m (\overline{[\lambda_{X = x_i}]} +
      \overline{[\lambda_{X = x_j}]})$\;
      \For{$x \in \im X$ {\rm \textbf{and}} $(y_1, \dots, y_n) \in \prod_{i = 1}^n \im
        Y_i$}{
        $p_x \gets \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n)$\;
        $\phi \gets \phi \cdot \left( p_x[\lambda_{X=x}] \cdot \prod_{i=1}^n
          [\lambda_{Y_i=y_i}] + \overline{[\lambda_{X=1}]} \cdot \prod_{i=1}^n
          [\lambda_{Y_i=y_i}] \right)$\;
      }
    }
  }
  \Return{$\phi$}\;
\end{algorithm}

\begin{lemma} \label{lemma:cpt}
  Let $X \in V$ be a random variable with $\mathrm{pa}(X) = \{ Y_1, \dots, Y_n
  \}$. Then the iteration of the main loop of the algorithm (when run on $X$)
  multiplies $\phi$ by a function $\mathrm{CPT}_X\colon 2^{E^*(X)} \to [0, 1]$
  such that
  \[
    \mathrm{CPT}_X(\tau) =
    \begin{cases}
      \Pr(X = x \mid Y_1 = y_1, \dots, Y_n = y_n) &
      \begin{aligned}
        &\text{if } \textstyle \tau = \lambda_{X=x} \land \bigwedge_{i=1}^n
        \lambda_{Y_i=y_i} \\
        &\text{for some } \textstyle (x, y_1, \dots, y_n) \in \im X \times
        \prod_{i=1}^n \im Y_i
        \end{aligned} \\
      0 & \text{otherwise.}
    \end{cases}
  \]
  for all $\tau \in 2^{E^*(X)}$.
\end{lemma}
\begin{proof}
\end{proof}

\begin{proposition} \label{lemma:full_distribution}
  $\prod_{X \in V} \mathrm{CPT}_X\colon 2^U \to [0, 1]$ represents the full
  probability distribution of the Bayesian network, i.e., if $V = \{ X_1, \dots,
  X_n\}$, then
  \[
    \left( \prod_{X \in V} \mathrm{CPT}_X \right)(\tau) =
    \begin{cases}
      \Pr(X_1 = x_1, \dots, X_n = x_n) & \text{if } \tau = \bigwedge_{i=1}^n
      \lambda_{X_i = x_i} \text{ for some } (x_1, \dots, x_n) \in \prod_{i=1}^n
      \im X_i \\
      0 & \text{otherwise,}
    \end{cases}
  \]
  for all $\tau \in 2^U$.
\end{proposition}
\begin{proof}
  We have that
  \[
    \left( \prod_{X \in V} \mathrm{CPT}_X \right)(\tau) = \prod_{X \in V}
    \mathrm{CPT}_X(\tau \cap E^*(X)).
  \]
  If $\mathrm{CPT}_X(\tau \cap E^*(X)) = 0$ for some $X \in V$, then $\left(
    \prod_{X \in V} \mathrm{CPT}_X \right)(\tau) = 0$. Otherwise, for each $X
  \in V$, $\tau \cap E^*(X)$ can be expressed as a conjunction of indicator
  variables. Moreover, those indicator variables must match, i.e.,
  \[
    \tau = \bigwedge_{X \in V} \lambda_{X = v_X}
  \]
  for some $(v_X)_{X \in V} \in \prod_{X \in V} \im X$. Then, by
  \cref{lemma:cpt},
  \[
    \left( \prod_{X \in V} \mathrm{CPT}_X \right)(\tau) = \prod_{X \in V} \Pr
    \left( X=v_X \;\middle|\; \bigwedge_{Y \in \mathrm{pa}(X)} Y=v_Y \right) =
    \Pr \left( \bigwedge_{X \in V} X=v_X \right),
  \]
  where the last step follows from the definition of a Bayesian network.
\end{proof}

\begin{theorem}
  Let $\phi\colon 2^U \to [0, 1]$ be a function generated by the algorithm. Then
  \[
    (\exists_U(\phi \cdot [\lambda_{X=x}]))(\emptyset) = \Pr(X = x).
  \]
\end{theorem}
\begin{proof}
  Let $V = \{ X, Y_1, \dots, Y_n \}$. Then
  \begin{align*}
    (\exists_U (\phi \cdot [\lambda_{X=x}]))(\emptyset) &= \sum_{\tau \in 2^U} (\phi \cdot [\lambda_{X=x}])(\tau) = \sum_{\lambda_{X=x} \in \tau \in 2^U} \phi(\tau) = \sum_{\lambda_{X=x} \in \tau \in 2^U} \left( \prod_{Y \in V} \mathrm{CPT}_Y \right)(\tau) \\
    &= \sum_{(y_1, \dots, y_n) \in \prod_{i=1}^n \im Y_i} \Pr(X = x, Y_1 = y_1, \dots, Y_n = y_n) = \Pr(X = x)
  \end{align*}
  by the following arguments:
  \begin{itemize}
  \item the proof of Theorem~1 in the ADDMC paper \cite{DBLP:conf/aaai/DudekPV20};
  \item if $\lambda_{X=x} \not\in \tau \in 2^U$, then $(\phi \cdot
    [\lambda_{X=x}])(\tau) = \phi(\tau) \cdot [\lambda_{X=x}](\tau \cap \{
    \lambda_{X=x} \}) = \phi(\tau) \cdot 0 = 0$;
  \item \cref{lemma:full_distribution};
  \item marginalisation of a probability distribution.
  \end{itemize}
\end{proof}

% Acknowledgments. This work has made use of the resources provided by the
% Edinburgh Compute and Data Facility (ECDF) (http://www.ecdf.ed.ac.uk/).
